Continuous gravitational waves in the lab: recovering audio signals with a table-top~optical microphone

### 2021-08-31 paper/ifo-complexAudio.tex
adding comments from Andrew

This analysis is indirectly related to the advanced signal processing techniques required to extract gravitational-wave signals. % citation?
In addition, the apparatus serves as an independent demonstration for a broader physics and engineering audience, particularly in undergraduate laboratories. 
We describe additional hardware components required for this demonstration in Section~\ref{sec:photodiode} and the initial results in Section~\ref{sec:initialResultsOpMic}. 
We consider a selection of filter techniques, details of which along with a summary of digital signal processing resources can be found in the Supplementary Material.
 % We also refer the reader to a selection of digital signal processing resources for further information. 
In Section~\ref{sec:opticalMicResults}, we present the two best-performing techniques from the Supplementary Material.

>>>>

%This analysis is indirectly related to the advanced signal processing techniques required to extract gravitational-wave signals. % citation?
The apparatus serves as an independent demonstration for a broader physics and engineering audience, particularly in undergraduate laboratories. 
We describe additional hardware components required for this demonstration in Section~\ref{sec:photodiode} and the initial results in Section~\ref{sec:initialResultsOpMic}. 
We consider a selection of filter techniques, details of which along with a summary of digital signal processing resources can be found in the Supplementary Material.
 % We also refer the reader to a selection of digital signal processing resources for further information. 
In Section~\ref{sec:opticalMicResults}, we present the two best-performing techniques from the Supplementary Material.

### 2021-08-31 paper/ifo-conclusions.tex
adding comments from Andrew

In this paper, we use a table-top Michelson interferometer as an analog to a gravitational-wave detector, demonstrating signal processing techniques used within the gravitational-wave community.
We explore the use of the interferometer as an optical microphone and consider a more general treatment of signal processing with complex audio signals, which can also serve as a distant analog to un-modeled gravitational-wave burst signals, e.g., from supernovae.
% justify ``un-modeled''? - james
The demonstration can be adapted for use in the undergraduate laboratory and as a tool for explaining gravitational-wave research to a wider, non-specialist audience.

>>>>

In this paper, we use a table-top Michelson interferometer as an analog to a gravitational-wave detector, demonstrating signal processing techniques used within the gravitational-wave community.
We explore the use of the interferometer as an optical microphone and consider a more general treatment of signal processing with complex audio signals, which can also serve as a distant analog to un-modeled gravitational-wave burst signals, e.g., from supernovae.
% justify ``un-modeled''? - james
The demonstration can be adapted for use in both the physics and engineering undergraduate laboratory, providing opportunities for cross-disciplinary teaching. 
Additionally, it can be used as a tool for explaining gravitational-wave research to a wider, non-specialist audience.

### 2021-08-31 paper/ifo-wanderingTone.tex
adding comments from Andrew

A continuous-wave signal may wander slowly (and randomly) in frequency over time, due to stochastic internal processes in the superfluid interior of isolated neutron stars, or variable accretion from a stellar companion for neutron stars in binaries as in LMXBs.
Although continuous gravitational waves are nearly monochromatic, the long observing times ($\sim 1\,{\rm yr}$) mean that searches are sensitive to very small frequency drifts. 
A typical observation involves $1 \times 10^{10}$ wave cycles, and a coherent search must track the phase to better than half a cycle over the full observation. 
Here we consider the audio analog of a tone that wanders in frequency.

A Fourier transform applied to the whole dataset (as in Section~\ref{sec:single_tone}) is not well suited to the wandering frequency case as the signal is spread across multiple frequency bins. 
In continuous-wave searches, the wandering frequency problem is solved using the Viterbi algorithm~\cite{Viterbi:1967}, which can track the signal's frequency over time.
The analysis described and presented in this section is directly inspired by real continuous-wave search methods~\cite{SuvorovaEtAl:2017}. 
In Section~\ref{sec:realCWSearches}, we briefly review the analysis method used by LIGO and Virgo; in Section~\ref{sec:viterbi}, we describe the methods as applied in this work; and in Section~\ref{sec:wanderingResults}, we show results for recovering a wandering frequency signal using the table-top interferometer.

@@

At the first stage of the analysis, we define the probability of the system having frequency $f_j$ at the initial time $t_0$ to be equal to the (normalized) detection statistic at that state (i.e.\  ${\rm Pr}[f(t_0) = f_j] = F[t_0,f_j]$).
A specific path (which may not be optimal) is written as $f(t_0), f(t_1),\mathellipsis, f(t_n)$. 
The probability of a specific path given the data is 
\begin{eqnarray}
{\rm Pr}[ f(t_0), f(t_1),&&\mathellipsis, f(t_n) | {\rm data}] \nonumber\\
          && =F[t_n,f(t_n)] A[f(t_n),f(t_n-1)] \nonumber \\
          && \times\mathellipsis\times A[f(t_1),f(t_0)] F[t_0,f(t_0)] \,,
\label{eqn:pathprobability}
\end{eqnarray}
The path $f^\ast(t_0),\mathellipsis,f^\ast(t_n)$ that maximises ${\rm Pr}[ f(t_0), f(t_1),\mathellipsis, f(t_n) | {\rm data}]$ is the optimal path terminating in the frequency bin $f^\ast(t_n)$. 
We note that the left-hand-side and right-hand-side of Eqn.~\ref{eqn:pathprobability} are both evaluated for a specific path and then we maximize over all such paths to find the optimal Viterbi path.

>>>>

A continuous-wave signal may wander slowly (and randomly) in frequency over time, due to stochastic internal processes in the superfluid interior of isolated neutron stars, or variable accretion from a stellar companion for neutron stars in binaries as in LMXBs (see Ref.~\cite{SuvorovaEtAl:2017} and references therein).
Although continuous gravitational waves are nearly monochromatic, the long observing times ($\sim 1\,{\rm yr}$) mean that searches are sensitive to very small frequency drifts. 
A typical observation involves $1 \times 10^{10}$ wave cycles, and a coherent search must track the phase to better than half a cycle over the full observation. 
Here we consider the audio analog of a tone that wanders in frequency.

A Fourier transform applied to the whole dataset (as in Section~\ref{sec:single_tone}) is not well suited to the wandering frequency case as the signal is spread across multiple frequency bins. 
In continuous-wave searches, the wandering frequency problem is solved using the Viterbi algorithm~\cite{Viterbi:1967}, which can track the signal's frequency over time.
The analysis described and presented in this section is directly inspired by real continuous-wave search methods~\cite{SuvorovaEtAl:2017} yet is pitched at a level appropriate for an undergraduate laboratory setting. 
In Section~\ref{sec:realCWSearches}, we briefly review the analysis method used by LIGO and Virgo; in Section~\ref{sec:viterbi}, we describe the methods as applied in this work; and in Section~\ref{sec:wanderingResults}, we show results for recovering a wandering frequency signal using the table-top interferometer.

@@

At the first stage of the analysis, we define the probability of the system having frequency $f_j$ at the initial time $t_0$ to be equal to the (normalized) detection statistic at that state (i.e.\  ${\rm Pr}[f(t_0) = f_j] = F[t_0,f_j]$).
A specific path (which may not be optimal) is written as $f(t_0), f(t_1),\mathellipsis, f(t_n)$. 
The probability of a specific path given the data is 
\begin{eqnarray}
{\rm Pr}[ f(t_0), f(t_1),&&\mathellipsis, f(t_n) | {\rm data}] \nonumber\\
          && =F[t_n,f(t_n)] A[f(t_n),f(t_{n-1})] \nonumber \\
          && \times\mathellipsis\times A[f(t_1),f(t_0)] F[t_0,f(t_0)] \,,
\label{eqn:pathprobability}
\end{eqnarray}
The path $f^\ast(t_0),\mathellipsis,f^\ast(t_n)$ that maximises ${\rm Pr}[ f(t_0), f(t_1),\mathellipsis, f(t_n) | {\rm data}]$ is the optimal path terminating in the frequency bin $f^\ast(t_n)$. 
We note that the left-hand-side and right-hand-side of Eqn.~\ref{eqn:pathprobability} are both evaluated for a specific path and then we maximize over all such paths to find the optimal Viterbi path.
